---
title: "GAM N-mixtures"
author: "Nicholas Clark"
date: "2024-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example 1
First we will use a simple simulation with multiple replicates at each timepoint to ensure that we can recover important temporal variation in latent abundance. The simulation produces observations for 6 timepoints with 5 replications per timepoint for 2 species, each with different temporal trends and detection probabilities
```{r include = FALSE}
library(mvgam)
library(ggplot2)
library(dplyr)
```

```{r eval = FALSE}
library(mvgam)
library(ggplot2)
library(dplyr)
```

```{r}
set.seed(999)
data.frame(site = 1,
           trend = 1,
           replicate = rep(1:5, 6),
           time = sort(rep(1:6, 5)),
           species = 'sp_1',
           truth = c(rep(28, 5),
                     rep(26, 5),
                     rep(23, 5),
                     rep(16, 5),
                     rep(14, 5),
                     rep(14, 5)),
           obs = c(rbinom(5, 28, 0.7),
                   rbinom(5, 26, 0.7),
                   rbinom(5, 23, 0.7),
                   rbinom(5, 15, 0.7),
                   rbinom(5, 14, 0.7),
                   rbinom(5, 14, 0.7))) %>%
  dplyr::mutate(series = paste0('site_', site,
                                '_', species,
                                '_rep_', replicate),
                time = as.numeric(time),
                cap = 50) %>%
  dplyr::select(- replicate) -> testdat

# Add another species
testdat = testdat %>%
  dplyr::bind_rows(data.frame(site = 1,
                              trend = 2,
                              replicate = rep(1:5, 6),
                              time = sort(rep(1:6, 5)),
                              species = 'sp_2',
                              truth = c(rep(4, 5),
                                        rep(7, 5),
                                        rep(15, 5),
                                        rep(16, 5),
                                        rep(19, 5),
                                        rep(18, 5)),
                              obs = c(rbinom(5, 4, 0.45),
                                      rbinom(5, 7, 0.45),
                                      rbinom(5, 15, 0.45),
                                      rbinom(5, 16, 0.45),
                                      rbinom(5, 19, 0.45),
                                      rbinom(5, 18, 0.45))) %>%
                     dplyr::mutate(series = paste0('site_', site,
                                                   '_', species,
                                                   '_rep_', replicate),
                                   time = as.numeric(time),
                                   cap = 30) %>%
                     dplyr::select(- replicate))
```


Add factor variables for indexing species and series IDs
```{r}
testdat$species <- as.factor(testdat$species)
testdat$series <- factor(testdat$series,
                         levels = unique(testdat$series))
```


The `trend_map` object operates by species here to state how the replicate observations need to be linked to the latent abundance processes
```{r}
trend_map <- testdat %>%
  dplyr::select(trend, series) %>%
  dplyr::distinct()
trend_map
```

Fit a model that allows species to have different detection probabilities
and different temporal trends (suppressing the output as the model runs to avoid seeing all the `Cmdstan` messaging)
```{r include = FALSE}
mod <- mvgam(
  # the observation formula sets up linear predictors for
  # detection probability
  formula = obs ~ species - 1,
  
  # the trend_formula sets up the linear predictors for 
  # the latent abundance processes 
  trend_formula = ~ s(time, by = trend, k = 4) +
        species,
  trend_map = trend_map,
  family = nmix(),
  data = testdat,
  # priors can be set using {brms} functionality
  priors = c(prior(std_normal(), class = b),
             prior(normal(1, 1.5), class = Intercept_trend)))
```

```{r eval = FALSE}
mod <- mvgam(
  # the observation formula sets up linear predictors for
  # detection probability
  formula = obs ~ species - 1,
  
  # the trend_formula sets up the linear predictors for 
  # the latent abundance processes 
  trend_formula = ~ s(time, by = species, k = 4) +
        species,
  trend_map = trend_map,
  family = nmix(),
  data = testdat,
  # priors can be set using {brms} functionality
  priors = c(prior(std_normal(), class = b),
             prior(normal(1, 1.5), class = Intercept_trend)))
```

View the automatically-generated `Stan` code to get a sense of how the marginalization over latent N works
```{r}
code(mod)
```

Summaries of key parameters
```{r}
summary(mod)
```

`{loo}` functionality works for all `{mvgam}` models to aid in model comparison / selection
```{r}
loo(mod)
```

Traceplots and all other types of MCMC diagnostics that are supported by `{bayesplot}` can be used to assess model failures or nonidentifiabilities. Here are traces of smoothing parameters
```{r}
mcmc_plot(mod, type = 'trace')
```

Plot the smooths of time from each species' latent abundance process (on the log scale)
```{r}
plot(mod, type = 'smooths', trend_effects = TRUE)
```

`{marginaleffects}` support allows for more useful prediction-based interrogations on different scales (i.e. link, response, detection or latent_N). Here are the estimated detection probabilities per species
```{r}
plot_predictions(mod, condition = list('species',
                                       cap = 10),
                 type = 'detection') +
  ylab('Pr(detection)') +
  ylim(c(0, 1)) +
  theme_classic() +
  theme(legend.position = 'none')
```

The model has automatically generated predictions of the latent abundance that are conditional on the observations. We can extract these and produce decent plots using a small function
```{r}
hc <- hindcast(mod, type = 'latent_N')

plot_latentN = function(hindcasts, data, species = 'sp_1'){
  # Grab the first replicate that represents this series
  series <- as.numeric(unique(data %>%
    dplyr::filter(species == !!species) %>%
    dplyr::pull(series))[1])
  truths <- data %>%
    dplyr::arrange(time, series) %>%
    dplyr::filter(series == !!levels(data$series)[series]) %>%
    dplyr::pull(truth)

  plot(hc,
       ylim = c(max(0, min(truths) - 5), max(truths) + 5),
       ylab = paste0('Latent abundance for ', species),
       series = series)

  points(truths, pch = 16, cex = 0.8, col = 'white')
  points(truths, pch = 16, cex = 0.65)
}
```

Latent abundance plots vs the simulated truths for each species show that low detection probabilities (like for species 2) make it difficult to accurately estimate latent abundance
```{r}
plot_latentN(hc, testdat, species = 'sp_1')
plot_latentN(hc, testdat, species = 'sp_2')
```

# Example 2
Now another example with a larger dataset. We will use data from Jeff Doser's simulation example from the `{spAbundance}` package:
[https://www.jeffdoser.com/files/spabundance-web/articles/nmixturemodels](https://www.jeffdoser.com/files/spabundance-web/articles/nmixturemodels). The simulated data include one continuous site-level covariate, one factor site-level covariate and two continuous sample-level covariates.
  
Download the data and grab observations / covariate measurements for one species. `NA`s allowed in responses but not in covariates
```{r}
load(url('https://github.com/doserjef/spAbundance/raw/main/data/dataNMixSim.rda'))
data.one.sp <- dataNMixSim
data.one.sp$y <- data.one.sp$y[1, , ]
abund.cov <- dataNMixSim$abund.covs[, 1]
abund.factor <- as.factor(dataNMixSim$abund.covs[, 2])
det.cov <- dataNMixSim$det.covs$det.cov.1[,]
det.cov[is.na(det.cov)] <- rnorm(length(which(is.na(det.cov))))
det.cov2 <- dataNMixSim$det.covs$det.cov.2
det.cov2[is.na(det.cov2)] <- rnorm(length(which(is.na(det.cov2))))
```

Wrangle into the appropriate 'long' data format, adding indicators of `time` and `series` for working in `{mvgam}`. We also add a `cap` variable to represent the maximum latent N to marginalize over for each observation
```{r}
mod_data <- do.call(rbind,
                    lapply(1:NROW(data.one.sp$y), function(x){
                      data.frame(y = data.one.sp$y[x,],
                                 abund_cov = abund.cov[x],
                                 abund_fac = abund.factor[x],
                                 det_cov = det.cov[x,],
                                 det_cov2 = det.cov2[x,],
                                 time = 1:NCOL(data.one.sp$y),
                                 site = paste0('site', x))
                    })) %>%
  dplyr::mutate(series = as.factor(paste0(site, '_', time)),
                site = factor(site, levels = unique(site))) %>%
  dplyr::mutate(time = 1,
                cap = max(data.one.sp$y, na.rm = TRUE) + 8)
```

Data features (225 sites, three replicates per site, though some observations are missing)
```{r}
NROW(mod_data)
dplyr::glimpse(mod_data)
head(mod_data)
```

The final step for data preparation is the `trend_map`, which sets up the mapping between observation replicates and the latent abundance models. Each site * species combination evolves as a latent 'factor', and the observation replicates can share these latent factors appropriately
```{r}
trend_map <- mod_data %>%
  dplyr::mutate(trend = as.numeric(site)) %>%
  dplyr::select(series, trend)
dplyr::glimpse(trend_map)
head(trend_map)
```

Fit a model using meanfield variational inference. Use penalized splines for each of the covariate effects to detect possible nonlinear associations. These linear predictors can easily handle multidimensional penalized smooths, time-varying or spatially-varying effects, Gaussian Process effects, random slopes and intercepts, or monotonic smooths. I'm purposely not suppressing the model output here so we can see how long it takes to fit
```{r}
mod <- mvgam(
  # effects of covariates on detection probability;
  # here we use penalized splines for both
  formula = y ~ s(det_cov, k = 4) + s(det_cov2, k = 4),
  # effects of the covariates on latent abundance;
  # here we use a penalized spline for the continuous and
  # hierarchical intercepts for the factor
  trend_formula = ~ s(abund_cov, k = 4) +
    s(abund_fac, bs = 're'),
  # linking multiple observations to each site
  trend_map = trend_map,
  family = nmix(),
  trend_model = 'None',
  data = mod_data,
  # standard normal priors on key regression params
  priors = c(prior(std_normal(), class = 'b'),
             prior(std_normal(), class = 'Intercept'),
             prior(std_normal(), class = 'Intercept_trend')),
  algorithm = 'meanfield',
  samples = 1000)
```

Inspect the model summary but don't bother looking at estimates for all individual spline coefficients
```{r}
summary(mod, include_betas = FALSE)
```

Use `{marginaleffects}` support for interrogating the model with more targeted predictions. First, inspect the estimated average detection probability
```{r}
avg_predictions(mod, type = 'detection')
```

Next investigate estimated effects of covariates on latent abundance
```{r}
abund_plots <- plot(conditional_effects(mod,
                                        type = 'link',
                                        effects = c('abund_cov',
                                                    'abund_fac')),
                    plot = FALSE)
```

The effect of the continuous covariate on expected latent abundance
```{r}
abund_plots[[1]] +
  ylab('Expected latent abundance')
```

The effect of the factor covariate on expected latent abundance, estimated as a hierarchical random effect
```{r}
abund_plots[[2]] +
  ylab('Expected latent abundance')
```

Investigate estimated effects of covariates on detection probability
```{r}
det_plots <- plot(conditional_effects(mod,
                                      type = 'detection',
                                      effects = c('det_cov',
                                                  'det_cov2')),
                  plot = FALSE)
```

The multidimensional smooth is estimated to be somewhat nonlinear on the logit scale according to the model summary (based on it's approximate significance). But inspecting conditional effects of each covariate on the probability scale is more intuitive and reportable
```{r}
det_plots[[1]] +
  ylab('Pr(detection)')
det_plots[[2]] +
  ylab('Pr(detection)')
```

More targeted predictions are also easy with `{marginaleffects}` support: How does detection probability change as we change both detection covariates?
```{r}
plot_predictions(mod, type = 'detection',
                 condition = c('det_cov',
                               'det_cov2')) +
  theme_classic() +
  ylab('Pr(detection)')
```
